Dice Enterprise

### 2019-01-15 inst/doc/DiceEnterprise.Rmd
Update vignettes

The `DiceEnterprise` package provides an automatic way to construct a Bernoulli Factory/Dice Enterprise algorithm for rational functions between simplices $f: \Delta^m \rightarrow \Delta^v$. That is, given an $m$-sided die, a $v$ sided-one is produced where the probability of each of the $v$ faces is a rational function. More formally, $f(\boldsymbol{p})$ is of the following form:
$$
f(\boldsymbol{p}) = f(p_1,\ldots,p_m) = \frac{1}{C(\boldsymbol{p})}\left(G_1(\boldsymbol{p},\ldots,G_v(\boldsymbol{p})\right)
$$
where each $G_i(\boldsymbol{p})$ and $C(\boldsymbol{p})$ are polynomials with real coefficients.

@@

The user can define a function that tosses a $p$-coin and returns $1$ with probability $p$ and 2 with probability $1-p$ (notice that **it should not return 1 or 0**). The function should have as input the number `n` of required tosses.  For instance the following function tosses a $\frac{3}{4}$-coin
```{r toss_coin}
toss_coin <- function(n) {
  sample(1:2, size = n, replace = TRUE, prob = c(3/4,1/4))
}
```

@@

### Example: Logistic Bernoulli Factory

Consider the function as in Huber
$$
f(p) = \frac{Cp}{1+Cp}
$$
Huber proposes a very clever algorithm that makes use of a thinned Poisson point process and requires on average $C/(1+Cp)$ tosses. If my calculations are correct, the Dice Enterprise can construct a Bernoulli Factory that albeit being always worse, requires at most one extra toss on average. Let's check this:

```{r bf_logistic}
toss.coin.logistic <- function(n,p) {
  sample(1:2, size = n, replace = TRUE, prob = c(p,1-p))
}
C <- 10
p <- 0.01
size_sample <- 10000

bf_logistic <- BernoulliFactory$new(f_1 = list(C,1), f_2 = list(1,0))
res <- bf_logistic$sample(n = size_sample, roll.fun = toss.coin.logistic, num_cores = 2, verbose = TRUE, p=p)

mean_Huber <- C/(1+C*p)
print(mean_Huber) #Mean of Huber Algorithm
print((1+C)/(1+C*p))  #Mean of Dice Enterprise
print(mean(res[[2]]))

#CHECK IF IT IS CORRECT
table(res[[1]])/size_sample
print(c(C*p/(1+C*p), 1/(1+C*p)))

```
```{r bf_Huber_A, echo=FALSE,eval=FALSE}
#Algorithm A of Huber for function 
#f(p) = Cp(1-(Cp)^(m-1))/(1-(Cp)^m) where m=5
m <- 5
bf_A <- BernoulliFactory$new(f_1 = list(coeff = 10^seq(m,2), power = seq(m-1,1)),
                             f_2 = list(coeff = c(10), power = c(0)))

size_sample <- 10000
res_A <- bf_A$sample(n = size_sample, roll.fun = toss.coin.logistic, num_cores = 2, verbose = TRUE, p=p)

mean_Huber_A <- C*(m-1)
print(mean_Huber_A) #Mean of Huber Algorithm
print(mean(res_A[[2]])) #Empirical mean of dice enterprise

#Check correctness
table(res_A[[1]])/size_sample
print(bf_A$evaluate(p))
```

## Dice Enterprise
The procedure is similar to the previous. Assume that the given function $f(\boldsymbol{p}): \Delta^m \rightarrow \Delta^v$ is 
$$
f(p_1,p_2,p_3) \propto (p_1p_2^2+\sqrt{2}p_2^2p_3^2, 4p_1^4p_2^7+1/2+3p_1p_2^2p_3^3,7p_1p_2^3p_3^4+2p_3^2)
$$
The user needs to define a list that specify the polymomial $f(\boldsymbol{p})$. Each element corresponds to $f_i(\boldsymbol{p})$ and each of this function is determined by a list itself containing two elements:
1. A vector of the coefficients (as in the Bernoulli Factory case)
2. A matrix or a vector of string defining the power of the variables. If a matrix is given, it has to have $m$ columns and each row defines the power of $p_1,p_2,\ldots,p_m$ respectively. For instance $p_1p_2^2+\sqrt{2}p_2^2p_3^2$ is translated into the matrix 
$$
\begin{bmatrix}
1 & 2 & 0 \\
0 & 2 & 2
\end{bmatrix}
$$
Alternatively, the same powers can be defined via a vector of strings. The powers of the previous polynomial $p_1p_2^2+\sqrt{2}p_2^2p_3^2$ can also be defined as `c("120","022")`. Notice that this method works **only** if there are no powers greater than 10.

@@

```{r example_2_paper}
de_ex2 <- DiceEnterprise$new(G=list(
  list(sqrt(2),"300"),
  list(1,"201"),
  list(1/4,"120"),
  list(2,"111"),
  list(1/2,"102"),
  list(3/4,"021")
))
#Define the original die
true_prob_original_die <- c(1/5,1/4,11/20)
roll_die <- function(n) {
  sample(1:3, size = n, replace = TRUE, prob = true_prob_original_die)
} 
#Get a sample of size 1000 from the multivariate ladder
#and plot the estimates with confidence intervals
sample_ex2 <- de_ex2$sample(n=1000, roll.fun = roll_die)
plot.confidence.interval(sample_ex2,de_ex2$evaluate(true_prob_original_die))

@@

Assume that instead of an $m$-sided die, $m$ independent coins are given, each one with its own probability of landing heads and denoted by $p_i$. In general, $(p_1,\ldots,p_m) \not\in \Delta^m$. Consider as an example, that we have access to 3 independent coins of biases $(0.4,0.7,0.55)$ and wish to sample from the distribution $f(\boldsymbol{p}) = \frac{1}{p_1+p_2+p_3}\left(p_1,p_2,p_3\right) = (8/33,14/33,1/3)$. There are several ways to define a die whose probabilites are a function of $(p_1,\ldots,p_m)$. The package provides support for two different definitions:

@@

>>>>

The `DiceEnterprise` package provides an automatic way to construct a Bernoulli Factory/Dice Enterprise algorithm for rational functions between simplices $f: \Delta^m \rightarrow \Delta^v$. That is, given an $m$-sided die, a $v$ sided-one is produced where the probability of each of the $v$ faces is a rational function. More formally, $f(\boldsymbol{p})$ is of the following form:
$$
f(\boldsymbol{p}) = f(p_1,\ldots,p_m) = \frac{1}{C(\boldsymbol{p})}\left(G_1(\boldsymbol{p}),\ldots,G_v(\boldsymbol{p})\right)
$$
where each $G_i(\boldsymbol{p})$ and $C(\boldsymbol{p})$ are polynomials with real coefficients.

@@

The user can define a function that tosses a $p$-coin and returns $1$ with probability $p$ and 2 with probability $1-p$ (notice that it **must** return 1 and 2). The function should have as input the number `n` of required tosses.  For instance the following function tosses a $\frac{3}{4}$-coin
```{r toss_coin}
toss_coin <- function(n) {
  sample(1:2, size = n, replace = TRUE, prob = c(3/4,1/4))
}
```

@@

When the function reaches close 1 or 0 within the interval $p \in (0,1)$, the construction of the ladder is computationally difficult and a runtime error may occur. Take for instance $f(p) = (x-1/2)^2 + 0.999$. The user can increase the timeout threshold by specifying the parameter threshold when constructing the BernoulliFactory object (default 100 iterations).

```{r bf_threshold, error=TRUE}
bf_error <- BernoulliFactory$new(f_1 = list(c(-1,1,0.749),c(2,1,0)), 
                                 f_2 = list(c(1,-1,0.251),c(2,1,0)))
bf_no_error <- BernoulliFactory$new(f_1 = list(c(-1,1,0.749),c(2,1,0)), 
                                 f_2 = list(c(1,-1,0.251),c(2,1,0)),
                                 threshold = 500)
```

<!--
### Alternative sampling algorithm

For Bernoulli Factory type problems (where $m = 2$) it is possible to use an alternative algorithm to get a sample based on acceptance-rejection.

```{r bf_AR}
fp_tosses_AR <- bf$sample.AR(n = 1000, roll.fun = toss_coin)
plot.confidence.interval(fp_tosses_AR,bf$evaluate(3/4))
```
By setting the option `verbose = TRUE`, the produced output is a list that contains also further information. In particular the empirical number of tosses required by the algorithm can be computed.
```{r bf_AR_verbose1}
fp_tosses_AR2 <- bf$sample.AR(n = 1000, roll.fun = toss_coin, verbose = TRUE)
print(paste0("Empirical number of tosses required: ",mean(fp_tosses_AR2$empirical_tosses)))
```

If instead of a function that tosses the original coin, the true value of $p$ is passed to the function and `verbose = TRUE`, further information are available. In particular, it is possible to compute the exact expected number of tosses required.

```{r bf_AR_verbose2}
fp_tosses_AR3 <- bf$sample.AR(n = 1000, true_p = c(3/4,1/4), verbose = TRUE)
print(paste0("Theoretical number of tosses required: ",mean(fp_tosses_AR3$theor_tosses)))
```
-->

## Dice Enterprise
The procedure is similar to the previous. Assume that the given function $f(\boldsymbol{p}): \Delta^m \rightarrow \Delta^v$ is 
$$
f(p_1,p_2,p_3) \propto (p_1p_2^2+\sqrt{2}p_2^2p_3^2, 4p_1^4p_2^7+1/2+3p_1p_2^2p_3^3,7p_1p_2^3p_3^4+2p_3^2)
$$
The user needs to define a list that specify the polymomial $f(\boldsymbol{p})$. Each element corresponds to $f_i(\boldsymbol{p})$ and each of this function is determined by a list itself containing two elements:

1. A vector of the coefficients (as in the Bernoulli Factory case)
2. A matrix or a vector of string defining the power of the variables. If a matrix is given, it has to have $m$ columns and each row defines the power of $p_1,p_2,\ldots,p_m$ respectively.
For instance $p_1p_2^2+\sqrt{2}p_2^2p_3^2$ is translated into the matrix 
$$
\begin{bmatrix}
1 & 2 & 0 \\
0 & 2 & 2
\end{bmatrix}
$$
Alternatively, the same powers can be defined via a vector of strings. The powers of the previous polynomial $p_1p_2^2+\sqrt{2}p_2^2p_3^2$ can also be defined as `c("120","022")`. Notice that this method works **only** if there are no powers greater than 10.

@@

```{r example_2_paper}
de_ex2 <- DiceEnterprise$new(G=list(
  list(sqrt(2),"300"),
  list(1,"201"),
  list(1/4,"120"),
  list(2,"111"),
  list(1/2,"102"),
  list(3/4,"021")
))
#Define the original die
true_prob_original_die <- c(1/5,1/4,11/20)
roll_die <- function(n) {
  sample(1:3, size = n, replace = TRUE, prob = true_prob_original_die)
} 
#Get a sample of size 1000 from the multivariate ladder
#and plot the estimates with confidence intervals
set.seed(17)
sample_ex2 <- de_ex2$sample(n=1000, roll.fun = roll_die, verbose = TRUE)
print(paste0("Average number of rolls required: ", mean(sample_ex2[[2]])))
plot.confidence.interval(sample_ex2[[1]],de_ex2$evaluate(true_prob_original_die))

@@

Assume that instead of an $m$-sided die, $m$ independent coins are given, each one with its own probability of landing heads and denoted by $p_i$. In general, $(p_1,\ldots,p_m) \not\in \Delta^m$. Consider as an example, that we have access to 3 independent coins of biases $(0.4,0.7,0.55)$ and wish to sample from the distribution $f(\boldsymbol{p}) = \frac{1}{p_1+p_2+p_3}\left(p_1,p_2,p_3\right) = (8/33,14/33,1/3)$. There are several ways to define a die whose probabilites are a function of $(p_1,\ldots,p_m)$. The package provides native support for three different definitions:

@@

```{r bug_to_correct, error=TRUE, echo = FALSE}
#(0.2501 - 1 x + 1 x^2)/((0.3 - 1 x + 1 x^2)+(0.2501 - 1 x + 1 x^2))
# bf_bug <- BernoulliFactory$new(
#   f_1 = list(c(0.2501,-1,1), c(0,1,2)),
#   f_2 = list(c(0.3,-1,1), c(0,1,2)),
# )
#(2-5x+5x^2)/(10x+10)
bf_bug <- DiceEnterprise$new(G = list(
  list(c(2,-5,5),c("00","10","20")),
  list(c(10,10),c("10","00"))
))
```
